---
title: "simulation functions"
author: "Lisa Rosenthal"
date: "12/4/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Functions for mesocosm simulation
Run in this order...

###STARTING DISTRIBUTIONS
1. species abundances: getabund2(n=100, mu=1) (need to run getabund())

###SIMULATION
1. format.comp(), 
2. HPraster(),  
3. simulate(h=format.comp, HP = HPraster, fun = f), 
4. f(), 
5. animate(HP = HPraster, v = simulate)

###RESPONSE VARIABLES
1. R1. response(sim, HP), 
2. response.sum = (response)


####load libraries and data management functions
```{r cars}

#libraries and display
library(raster)
library(dplyr)
library(ggplot2)
library(reshape2)
library(RColorBrewer)
#display.brewer.all()
pal <-rev(brewer.pal(6, "YlOrRd"))  

#data management functions
extract_after <-  function(char, vector){
    pattern <- paste0(".*", char)
    sub(pattern, '', vector)#after chacter
  }
```


###SPECIES DISTRIBUTIONS
lognormal distribution of species abundances
important for knowing abundances of species at each richness level.

you have 6 species. get proportion of individuals per species and standardize to a certain number of total individuals.
```{r}

getabund <- function(r, n=100, mu=1){ 
  #r=richness, n=total individuals, mu=mean of log distribution
  
#get counts of each species
  n1 = n+max(r) #buffering the number of species so there aren't errors due to rounding
  pool=dlnorm(1:6, mu) #distribution of each species
  abund <- pool[1:r]*n1/sum(pool[1:r]) #get abundances of each species for a richness level

#get actual population from the counts above 
  r2=1:length(r)
  pop <- rep(1:length(abund), abund)
  pop <- sort(sample(pop, n)) #bring the population back down to desired n. had issues with rounding.
  return(pop)
}

getabund2 <- function(n=100, mu=1){
  x=c(1,2,4,6)
  counts <- lapply(x, function(r) getabund(r, n, mu))
  counts2 <- unlist(counts)
  counts2 <- as.data.frame(matrix(counts2, ncol=length(x))) 
  names(counts2) <- c("R1", "R2", "R3", "R4")
  counts2
}
```


###SIMULATION
```{r}
############################################################
#1. format.comp()
############################################################
#make function to get host values and matrix of infection probabilities
format.comp <- function(compdf=comp2, plants=abund$R2, steps=20){
  #plants=vector of plant identies
  #steps=total time of simulation
  #compdf=dataframe of competency values over time. eg. comp2. column values are time, species, comp. length(time)=steps.
  
  Rename <-  function(values_to, index_from, from_column){
    result <- (values_to[match(from_column, index_from)])
    return(result)
  }
  
  #assign values to host
  s <- matrix(rep(plants, steps), ncol=steps) #species matrix across time
  s <- s[sample(nrow(s)),] #randomize cells
  
  #get matrix of competency values for each time step after exposure
  val <- lapply(1:steps, function(x) Rename(compdf[compdf$time==x,]$comp, compdf[compdf$time==x,]$species, from_column = s[,x]))
  h <- matrix(unlist(val), ncol = length(val), byrow = F)
  return(list("infection_prob"=h, "species_ident"=s))
}

############################################################
#2. HPraster()
############################################################
#set up pathogen and host raster grids
HPraster <- function(h, n=10, ncol=10, nrow=10){
  #h=output from format.comp()
  #n=number of inoculated plants
  #ncol/nrow dimensions of raster grid
  
  host <- path <- raster(ncol=ncol, nrow=nrow, crs="+proj=utm +zone=1 +datum=WGS84", xmn=0, xmx=10, ymn=0, ymx=10) #create a raster layer for the host and pathogen
  
  #host
  values(host) <- h[[2]][,1] # species identity, already randomized
  
  #inoculate with pathogen
  values(path) <- 0
  values(path) <- sample(c(rep(1, n), rep(0, ncell(path)-n)), replace = F)
  
  return(list("path"=path, "host"=host))
}

############################################################
#3. simulate()
############################################################
#simulation
simulate <- function(h, HP, steps, fun=f) {
  #setting up the raster grid
  h <- h[[1]]  
  path <- HP$path
  host <- HP$host
  v <- matrix(0, ncell(path), steps)
  v
  v[,1] <- values(path) #values of infection status
  a <- adjacent(host, 1:ncell(host), 4, sorted=TRUE)
  com.rates <- matrix(rep(h[,1], steps), ncol = steps) #probability of infection rates over time, aka competency. change after exposure
  t.exposed <- rep(NA, ncell(host))
  
  for (time in 1:(steps-1)) { #for every time interval
    v[,time+1] <- v[,time] #previous status continues to next time step
    
    #chunk for getting time since exposure for each cell
    for (i in 1:nrow(v)) { #for every cell
      if (v[i, time+1] == 1 ) next #skipping those already infected
      adj <- a[a[,1] == i, 2]	#find all your neighbors
      if (any(v[adj, time] > 0)) { #if any of the neighbors are infected
        if ( !is.na(t.exposed[i]) ) next #skipping those already exposed
        t.exposed[i] <- (time) #record first time exposed
      }
    }
    
    #chunk for turning susceptibles into infected
    for (i in 1:nrow(v)) { #for every cell
      if (v[i, time+1] == 1 ) next #skipping those already infected
      adj <- a[a[,1] == i, 2]	#find all your neighbors
      if (any(v[adj, time] > 0)) { #if any neighbors are infected
        com.rates[i, time] <- h[i, time-t.exposed[i]+1] #update probability of infection as a function of time after exposure
        v[i, time+1] <- fun(v[i, time], com.rates[i, time]) #become infected at some function of current health status and host competency
      } 
    }
  }
  list("values"=v, "Inf.prob"=com.rates, "t.exposed"=t.exposed)
}

############################################################
#4. f()
############################################################
#how to determine if cell is infected
f <- function(p, h) {
  if ( runif(1, 0.1, 1) <= h) {#youre infected if value is less than or equal your probability of getting infected (competency value). h=1 always infected. h=0 never infected.
    1
  } else {
    p #if not infected, remain the same as last time step, p
  }
}

############################################################
#5. animate()
############################################################
#for visualizing
animate <- function(HP, v, pause=0.25, col=pal) {
  v <- v[[1]]
  h <- HP$host
  p <- HP$path
  
  for (i in 1:ncol(v)) { #for every time step
    values(p) <- v[,i] #assign the values of matrix v to raster p at each time step
    plot(h, asp=NA, col=col, main=i)
    plot(p, legend=FALSE, asp=NA, col=scales::alpha(c(NA, "black"), .8), add=T) #plot it
    dev.flush() #not sure what this does
    Sys.sleep(pause) #give me .25 sec to see each frame
  }
}

```

###Response variables
I need reports on response variables for each simulation set

2 functions: 
1. `response`=RV for each time step and species within community, 
2. `response.sum` = summary response variables to compare across treatments
```{r}

response <- function(sim, HP){
#data=output from simulate; HP=output from HPraster
  data <- sim
#get data
dat <- data[[1]] #state of each cell by time: N x time
spp <- as.vector(values(HP$host)) #vector of species present in correct

#get table of frequency of each species 
fr <- as.data.frame(table(spp))
N <- cbind.data.frame("tot", as.numeric(sum(fr$Freq)))
names(N) <- names(fr) <- c("species", "n")
fr <- rbind(fr, N)

############################################################
#make tables of infecteds
############################################################
#for each time step, get sum of infecteds and cbind onto existing f dataframe
sp <- fr$species[1:length(fr$species)-1]
I <- (dat * spp)
frI <- fr
for(t in 1:ncol(I)){
  tmp <- sapply(sp, function(i) {
    sum(I[,t]==sp[i])
  })
  tmp[length(tmp)+1] <- sum(tmp)
  frI <- cbind(frI, tmp)
}
names(frI) <- c(names(fr), 1:ncol(I))

############################################################
#make tables of exposed
############################################################
#for each time step, get sum of exposed and cbind onto existing f dataframe

#find time infected for every cell
t.inf <- sapply(1:nrow(dat), function(i) which(dat[i, ]==1)[1])

#table of exposure
exposure <- data.frame(cell=1:length(data$t.exposed), exp.i=data$t.exposed, exp.f=t.inf-1)
#$exp.f==NA when never infected. $exp.i==NA when never exposed. NA's introduced when 1. never exposed, 2. inoculated, 3. exposed, but not infected
#1. never exposed
S <- is.na(exposure$exp.i) & is.na(exposure$exp.f)
#2. inoculated
I <- is.na(exposure$exp.i) & exposure$exp.f==0
#3. exposed, never infected
E <- exposure$exp.i>0 & is.na(exposure$exp.f)
#change values
exposure$exp.i[which(S)] <- 0
exposure$exp.i[which(I)] <- 0
exposure$exp.f[which(S)] <- 0
exposure$exp.f[which(I)] <- 0
exposure$exp.f[which(E)] <- ncol(dat)

for(i in 1:nrow(dat)){
  #change these values to NA, corresponding to time exposed
  dat[i, exposure$exp.i[i]:exposure$exp.f[i]] <- NA 
}

#make matrix for just exposed plants
e <- dat
e[!is.na(e)] <- 0
e[is.na(e)] <- 1

#get frequencies of exposed plants
E <- e * spp

frE <- fr
for(t in 1:ncol(E)){
  tmp <- sapply(sp, function(i) {
    sum(E[,t]==sp[i])
  })
  tmp[length(tmp)+1] <- sum(tmp)
  frE <- cbind(frE, E=tmp)
}
names(frE) <- c(names(fr), 1:ncol(E))


############################################################
#make table tall
############################################################
frI2 <- melt(frI, c("species", "n"), 3:ncol(frI), variable.name = "time", value.name = "n.I")
frE2 <- melt(frE, c("species", "n"), 3:ncol(frE), variable.name = "time", value.name = "n.E")
fr2 <- left_join(frI2, frE2, by=c("species", "time", "n")) %>% arrange( species) 

############################################################
#get n.S and %inf
############################################################
fr2$n.S <- fr2$n-fr2$n.I
fr2$pI <- fr2$n.I/fr2$n

############################################################
#get dI/dt
############################################################
#dI/dt = I[t+1]-I[t]
di <- function(t) fr2[t,]$n.I-fr2[t-1,]$n.I
fr2$dI <- c(NA, di(2:length(fr2$n.I))) 

#change all dI where time=1 to NA
fr2$dI[fr2$time==1] <- NA

#time needs to be numerical
fr2$time <- as.numeric(fr2$time)

return(fr2)
}

response.sum <- function(resp){
  #n.species, n.infected, perc.infected for each species
  RV1 <- resp %>% 
    filter(time==max(time)) %>% 
    group_by(species) %>% 
    summarize(n, n.I, pI)
  
  #dIdt reports
  RV2 <- resp %>% 
    group_by(species) %>% 
    filter(dI==max(dI, na.rm = T)) %>% 
    filter(!duplicated(dI)) %>% #remove rows with duplicated dI
    summarise(max.dI=dI, tmax.dI=time)
  
  left_join(RV1, RV2, by="species")
}
```

