---
title: "Fitting_2"
author: "Lisa Rosenthal"
date: "4/7/2020"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rstan)
library(rethinking)
library(tidyverse)
library(tidybayes)
library(abind)
library(shinystan)
rm(list=ls())

#convienence functions
headmat <- function(matrix, nrow=6, ncol=6) matrix[1:nrow, 1:ncol]

```

## Simple SIR model  

Let's say there are 3 types of plants: those that are inoculated/challenged (C), susceptible (S), and infected (I). We can observe the states of individuals $i=1...n$ at times $t=1...T_{f}$. The total number of individuals in each state at time $t$ are $C_t$, $S_t$, and $I_t$. Challenged individuals can become infected either through inoculum-plant transmission at a rate $\alpha$ or plant-plant transmission at a rate $\beta I$. Since we assume that inoculum can only infect individuals a short distance away (i.e. the "challenged" individuals), susceptible individuals can only be infected through plant-plant transmission.  

We need to model the transitions $C \rightarrow I$ and $S \rightarrow I$. The approximated probability of either a challenged or susceptible individual $i$ becoming infected at time $t$ conditional upon $I_{t-1}$, the infectious population from the previous time step, are  

\begin{align}
\text{P}(i \in I_t | i \in C_{t-1}, I_{t-1}) = 1 - \exp(-(\alpha + \beta I_{t-1}))\\
\text{P}(i \in I_t | i \in S_{t-1}, I_{t-1}) = 1 - \exp(-\beta I_{t-1}).
\end{align}

The likelihood of infection for each individual is a binomial process determined by the probabilities defined in (1) and (2).  

## Simulate data  

For now, I'm going to ignore heterogeneity in space and species, and assume transmission coefficients do not vary with time. 

```{r}
source('IBM/scripts/IBM_functions.R')

#starting parameters
pinoc <- .1 #percent inoculated
s <- 1 #number of spp
spp <- "sp1"
tfinal <- 6 #time steps
beta <- .05
alpha <- .1
r <- 2 #interplanting distance in cm
nreps <- 10

#simulate communities
grid <- replicate(nreps, sample_community(1, pinoc, r)) #function creates community in a 9.5in square tray, plants in hexagonal grid
N <- sapply(grid, length) #number of individuals

#plot_maps(grid[[1]])

#run simulation
f_sim1 <- function(grid, N, pinoc=pinoc, s=s, spp=spp, tfinal=tfinal, beta=beta, alpha=alpha, r=r){

  #keep track of all individuals' states
  agents <- grid@data
  states_matrix <- matrix(NA, nrow=N, ncol = tfinal)
  states_matrix[,1] <- as.character(agents$state0)
  
  #vectors denoting specific states
  challenged <- as.numeric(agents$state0=='C')
  susceptible <- 1-challenged
  infected <- rep(0, N)
  
  #Run the simulation
  for(t in 1:(tfinal-1)){ #at every time step
    
    #calculate probabilities of infections
    P_ci <- (1 - exp(-(alpha + beta*sum(infected))))*challenged
    P_si <- (1 - exp(-beta * sum(infected)))*susceptible
    
    #generate new infections
    CI <- rbinom(N, 1, P_ci)
    SI <- rbinom(N, 1, P_si)
    new_infections <- CI + SI
    
    #update states
    challenged <- challenged - CI
    susceptible <- susceptible - SI
    infected <- infected + new_infections
    states_matrix[,t+1][challenged==1] <- 'C'
    states_matrix[,t+1][susceptible==1] <- 'S'
    states_matrix[,t+1][infected==1] <- 'I'
  }
  
  return(states_matrix)
}

#do simulation multiple times
out <- NULL
for(i in 1:nreps){
  out[[i]] <- as.data.frame(f_sim1(grid = grid[[i]], N=N[i], pinoc=pinoc, s=s, spp=spp, tfinal=tfinal, beta=beta, alpha=alpha, r=r))
  out[[i]]$rep <- i
  out[[i]]$ID <- 1:N[i]
}
out <- bind_rows(out) 

#visualize
dat <- out %>% 
  pivot_longer(cols = starts_with("V"), names_to = 'time', values_to = 'state') %>% 
  mutate(time = as.numeric(gsub("V", '', time)), ID = as.factor(ID))

#for every individual
ggplot(filter(dat, rep==1), aes(time, ID, fill=state)) +
  geom_tile() +
  scale_fill_manual(values = c('C' = 'orange', 'S' = 'steelblue3', 'I' = 'indianred')) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())

#as a population
dat_pop <- dat %>% count(rep, time, state)
ggplot(dat_pop, aes(time, n, color=state, group=interaction(state, rep))) +
  geom_line(alpha=.5) +
  scale_color_manual(values = c('C' = 'orange', 'S' = 'steelblue3', 'I' = 'indianred'))

#plot_spread_map(grid, states_matrix, animate = T) #transmission is spatially random and fast

```


Generate two types of data: non-replicated and replicated. I started with writing a model with non-replicated, got it to mostly work. Now I'm working on adding complexity to the replicated data. 
```{r}

#non-replicated
outNR <- as.data.frame(f_sim1(grid = grid[[1]], N=N[1], pinoc=pinoc, s=s, spp=spp, tfinal=tfinal, beta=beta, alpha=alpha, r=r))
outNR$rep <- 1
outNR$ID <- 1:N[1]
dat <- outNR %>% 
  pivot_longer(cols = starts_with("V"), names_to = 'time', values_to = 'state') %>% 
  mutate(time = as.numeric(gsub("V", '', time)), ID = as.factor(ID))
dat_pop <- dat %>% count(rep, time, state)


#replicated
#put data into an array
nreps = 2
outR <- NULL
for(i in 1:nreps){
  outR[[i]] <- f_sim1(grid = grid[[i]], N=N[i], pinoc=pinoc, s=s, spp=spp, tfinal=tfinal, beta=beta, alpha=alpha, r=r)
}
outR <- abind::abind(outR, along = 3) #array of states.
```


## Fitting model in Stan  

I'm going to estimate alpha and beta for this simple disease model. This will act as the building block for more complex models. I'll show how to model it with aggregated data (binomial distribution) and disaggregated (bernoulli random variables). Modeling individuals independently with the bernoulli likelihood should make it easier to add more complexity in the future when all the individuals are different.  

#### Aggregated data modeled with binomial distribution:  
Code for non-replicated data. Runs really fast!
```{r}
binom_NR = 
"
data { 
  //declare variables
  int<lower=0> N; //number of individuals in each tray
  int<lower=0> t; //number of time points
  //states
  int<lower = 0, upper =N> C[t];
  int<lower = 0, upper =N> S[t]; 
  int<lower = 0, upper =N> I[t];
}
transformed data { 
  //Calculate transitions, based on change in states
  int<lower=0, upper = N> z_ci[t-1]; 
  int<lower=0, upper = N> z_si[t-1];
  
  //I think it should be for(i in 2:t) z_ci[2] = C[1] - C[2]?
    for(i in 1:(t-1)){
    z_ci[i] = C[i] - C[i+1]; //new infections of C
    z_si[i] = S[i] - S[i+1]; //new infections of S
    }
}
parameters {
  real<lower=0, upper = 1> alpha; //I think it's also between 0 and 1.
  real<lower=0, upper = 1> beta;
}

model {
  real p_ci; //prob. of C -> I
  real p_si; //prob. of S -> I
  
  //priors
  alpha ~ beta(1,5); //previously had at gamma(1,1)
  beta ~ beta(1,5);

 //generate number of infecteds from C and S at each time
 //C, S, and I are the number of individuals in that state at time t
 for(i in 1:(t-1)){
    p_ci = 1 - exp(-(alpha + beta*I[i]));
    z_ci[i] ~ binomial(C[i], p_ci); 
  //only define z_si when there are infections. distribution is degenerate and STAN has trouble
    if(I[i] > 0){
      p_si = 1 - exp(-(beta*I[i]));
      z_si[i] ~ binomial(S[i], p_si); 
    }
 }
}
"
binom_NR_mod = stan_model(model_code = binom_NR)


# format data for STAN
dat_NR <- dat_pop %>% 
  pivot_wider(names_from = state, values_from = n, values_fill = list(n=0)) %>% 
  dplyr::select(-time) %>% 
  as.list %>% 
  c(t = as.integer(tfinal)) %>% list_modify(N=N[1])

str(dat_NR) #make sure integers are classified as such

fit_binom_NR = sampling(binom_NR_mod,dat_NR, iter = 2000, chains = 3)

#check it out
precis(fit_binom_NR) %>% plot #covers the parameters. Interval of alpha is big. 
stan_dens(fit_binom_NR)
print(fit_binom_NR,digits = 3)
traceplot(fit_binom_NR)

```



#### Bernoulli distribution

Here's the code for non-replicated data. same results as binomial, just much slower. 
```{r}
#need to index data by time t AND individual N
bern_NR = 
"
data { 
  //declare variables
  int<lower=0> N; //number of individuals
  int<lower=0> t; //number of time points
  int<lower = 0, upper =N> C[N,t]; //states
  int<lower = 0, upper =N> S[N,t]; 
  int<lower = 0, upper =N> I[N,t];
}
transformed data { 
  //Calculate transitions, based on change in states
  int<lower=0, upper = N> z_ci[N,t-1]; 
  int<lower=0, upper = N> z_si[N,t-1];
  for(T in 1:(t-1)){
    for(i in 1:N){
      z_ci[i,T] = C[i,T] - C[i,T+1]; //new infections of C
      z_si[i,T] = S[i,T] - S[i,T+1]; //new infections of S
    }
  }
}
parameters {
  real<lower=0, upper = 1> alpha;
  real<lower=0, upper = 1> beta;
}

model {
  real p_ci;
  real p_si;
  
  //priors
  alpha ~ beta(1,5);
  beta ~ beta(1,5);

 //generate number of infecteds from C and S at each time
 //C, S, and I are the number of individuals in that state at time t
 for(T in 1:(t-1)){
  for(i in 1:N){
    if(C[i,T]){ //transition for challenged
      p_ci = 1 - exp(-(alpha + beta*sum(I[,T])));
      z_ci[i,T] ~ bernoulli(p_ci); 
    }
    if(sum(I[,T]) > 0 && S[i,T]){ //transition for susceptible
      p_si = 1 - exp(-beta*sum(I[,T]));
      z_si[i,T] ~ bernoulli(p_si); 
      }
    }
 }
}
"
mod_bern_NR = stan_model(model_code = bern_NR)

#format data for model
# format data for STAN
dat_NR2 <- list(
  N = nrow(outNR), 
  t = as.integer(tfinal),
  C = matrix(as.numeric(outNR[,1:tfinal] == "C"), ncol = tfinal),
  S = matrix(as.numeric(outNR[,1:tfinal] == "S"), ncol = tfinal),
  I = matrix(as.numeric(outNR[,1:tfinal] == "I"), ncol = tfinal)
)
str(dat_NR2) #make sure integers are classified as such



#run model
fit_bern_NR = sampling(mod_bern_NR,dat_NR2, iter = 1000, chains = 2, cores = 2)

#check it out
precis(fit_bern_NR) 
par(mfrow=c(1,2)) 
precis(fit_bern_NR) %>% plot; precis(fit_binom_NR) %>% plot #binomial and bernoulli the same. 
par(mfrow=c(1,1)) 
traceplot(fit_bern_NR)


my_sso <- launch_shinystan(fit_bern_NR)
```


Trying the bernoulli model with replicated data.
```{r}
# format data for STAN
dat_R1 <- list(
  N = as.integer(dim(outR)[1]),
  J = max(nreps),
  t = as.integer(tfinal),
  ID = as.integer(1:dim(outR)[3]),
  C = abind(lapply(1:nreps, function(i) matrix(as.numeric(outR[,,i] == "C"), ncol = tfinal)), along = 3),
  S = abind(lapply(1:nreps, function(i) matrix(as.numeric(outR[,,i] == "S"), ncol = tfinal)), along = 3),
  I = abind(lapply(1:nreps, function(i) matrix(as.numeric(outR[,,i] == "I"), ncol = tfinal)), along = 3)
)
str(dat_R1) #make sure integers are classified as such

mod_bern1 <- stan_model(file = "IBM/scripts/model_fitting/sir_bernoulli1.stan")
fit_bern1 <- sampling(mod_bern1, dat_R1, iter = 1000, chains = 2, cores = 2)

#alpha sd is waaaay higher than beta. But that might be that there are a lot fewer chalenged individuals to estimate that parameter. The estimates are ok though, albeit with big CI.
#having 2 seperate varying intercepts for C->I and S->I underpredicts alpha.
precis(fit_bern1, depth = 2)
plot(precis(fit_bern1, depth = 2, pars = c('alpha', 'beta', 'zz')))
precis(fit_bern1, pars = 'beta') 
traceplot(fit_bern1) #need to non center probably
post <- extract.samples(fit_bern1)
dens(post$beta)
sd(post$beta)


#2 independent random effects (gamma distributed) are added to P(C-I) and P(S-I). Took FOREVER.
mod_bern1_1 <- stan_model(file = "IBM/scripts/model_fitting/sir_bernoulli1-1.stan")
fit_bern1_1 <- sampling(mod_bern1_1, dat_R1, iter = 500, chains = 1, cores = 1)
precis(fit_bern1_1, depth = 2)
precis(fit_bern1_1, depth = 2, pars = c('alpha', 'beta')) %>% plot
traceplot(fit_bern1_1)


#try to fit where random effects are embedded within alpha and beta. individual estimates of alpha and beta look good, but the estimated mean too high and doesn't make sense.
mod_bern1_2 <- stan_model(file = "IBM/scripts/model_fitting/sir_bernoulli1-2.stan")
fit_bern1_2 <- sampling(mod_bern1_2, dat_R1, iter = 500, chains = 1, cores = 1)
precis(fit_bern1_2, depth = 2)
precis(fit_bern1_2, depth = 2, pars = c('alpha', 'beta')) %>% plot
traceplot(fit_bern1_2)

#try to fit where random effects multiply alpha and beta. really large intervals for both a and b.
mod_bern1_3 <- stan_model(file = "IBM/scripts/model_fitting/sir_bernoulli1-3.stan")
fit_bern1_3 <- sampling(mod_bern1_3, dat_R1, iter = 500, chains = 1, cores = 1)
precis(fit_bern1_3, depth = 2)
precis(fit_bern1_3, depth = 2, pars = c('alpha', 'beta')) %>% plot
traceplot(fit_bern1_3)


dens(rgamma2(10000, .04, .59))
rgamma2(100000, .2, .2) %>% dens
rgamma2(100000, .2, .2) %>% mean
rbeta(100000, 1, 5) %>% dens
```



All of the models above estimate alpha and beta within the ranges of what I used to simulate the data. That's a good sign. The model with replicated data has smaller intervals of alpha, which is what I'd expect. The aggregated data model is really fast, but unfortunately the disaggregated one isn't vectorized so it goes really slow.  

Below I try to add in more complexity by allowing 1) the transmission coefficients to vary with time, 2) species to vary in probability of infection given contact, and 3) contact networks to vary by the spatial arrangements of individuals. I will need to use the bernoulli model from here on out.  

## Simulate more heterogeneity  

- transmission coefs vary in time  
- transmission decays with distance . 
- species vary in suceptibility and infectivity (worked on last)  

```{r}

#NEW FUNCTIONS----

#beta will be a product of contact rate X probability of transmission given contact
#contact rate may vary with distance
dist_decay <- function(x, Sigma){
  #equation from Kleczkowski et al. (1997).  sigma scales the rate of the decay. Other parameters in the paper affect a constant, but it cancels out when I standardize it.
  #probably a way to add a gaussian distribution as the predictor instead. 
  exp(-Sigma*x^2)
}

#probability of transmission given contact varies with time
alpha_curve <- function(t, amp=.4, rate=1) amp*exp(-rate*t)
beta_curve <- function(t, amp=.2, rate=3, tq=11) {
  #amp=amplitude
  #rate=variance of curve. larger is less variance.
  #tq=time of peak
  amp*exp(-rate*(log(t/tq))^2)
}

#will assume different spatial structures. Contact may be equally mixed, be between nearest neighbor only, or decay with distance. 

dist_equal <- function(grid) matrix(1, nrow = length(grid), ncol = length(grid))#homogenously mixed, contacts all same
dist_NN <- function(grid) matrix(as.numeric(rgeos::gTouches(grid, byid = T)), nrow = length(grid), ncol = length(grid)) #nearest neighbor
dist_euc <- function(grid) as.matrix(dist(cbind(grid$x, grid$y)))*10 #euclidian distance
f_kernel <- function(dist_matrix, sigma){
  contact_rt <- dist_decay(dist_matrix, Sigma = sigma)
  diag(contact_rt) <- 0 #Self can't infect self.
  return(contact_rt)
}

#that introduces new parameters: amp_a, rate_a, amp_b, rate_b, tq, sigma. 
#When I model different species, I'm going to assume that species change just the amplitude. Possibly a restrictive assumption, but can't introduce too many parameters. 


#PARAMETERS----

#same parameters as above
pinoc <- .05 #percent inoculated
s <- 1 #number of spp
spp <- "sp1"
tfinal <- 20 #time steps
r <- 2

#new parameters (will be estimated in model fitting)
amp_a <- .6
rate_a <- .5
amp_b <- .1
rate_b <- 1
tq <- 10
sigma <- .001 #higher value -> less dispersal
pinoc <- .1

#simulate communities
grid <- replicate(nreps, sample_community(1, pinoc, r)) 
N <- sapply(grid, length)


#visualize transmission parameters
dist_matrix <- dist_euc(grid[[1]])
contact_mat <- f_kernel(dist_matrix, sigma = sigma) #contact matrix. for NN and mass action, contact matrix will be the same as the dist_matrix.

par(mfrow=c(1,2))
#alpha and beta
xtime <- seq(0,20, by=.1)
plot(xtime, alpha_curve(t=xtime, amp=amp_a, rate=rate_a), ylim=c(0,1), type='l', col='steelblue', lwd=2, xlab='days', ylab='transmission rate')
lines(xtime, beta_curve(t=xtime, amp = amp_b, rate=rate_b, tq=tq), type='l', col='indianred',lwd=2)
legend('topright', legend = c('alpha', 'beta'), fill = c('indianred', 'steelblue'))
#distance decay
xdist <- seq(0,50, by=1) #distance in mm
plot(xdist, dist_decay(xdist, Sigma = sigma), type="l", xlab='distance (mm)', ylab="P(contact)", lwd=2)
legend('topright', legend = c('distance decay'), fill = c('black'))
par(mfrow=c(1,1))



############################################################
#simulate
############################################################

fsim2 <- function(grid, N, sigma, tfinal, amp_a, rate_a, amp_b, rate_b){
 
  #contact matrix. for NN and mass action, contact matrix will be the same as the dist_matrix.
  dist_matrix <- dist_euc(grid)
  contact_mat <- f_kernel(dist_matrix, sigma) 
  
  #keep track of all individuals' states
  agents <- grid@data
  states_matrix <- matrix(NA, nrow=N, ncol = tfinal)
  states_matrix[,1] <- as.character(agents$state0)
  
  #vectors denoting specific states
  challenged <- as.numeric(agents$state0=='C')
  susceptible <- 1-challenged
  infected <- rep(0, N)
  
  #Run the simulation
  for(t in 1:(tfinal-1)){ 
    
    #calculate prob of transmission given contact
    alpha <- alpha_curve(t = t, amp = amp_a, rate = rate_a)
    beta <- beta_curve(t = t, amp = amp_b, rate = rate_b)
    
    #calculate probabilities of infections
    P_ci <- (1 - exp(-(alpha + beta*(contact_mat %*% infected))))*challenged
    P_si <- (1 - exp(-beta*(contact_mat %*% infected)))*susceptible
    
    #generate new infections
    CI <- rbinom(N, 1, P_ci)
    SI <- rbinom(N, 1, P_si)
    new_infections <- CI + SI
    
    #update states
    challenged <- challenged - CI
    susceptible <- susceptible - SI
    infected <- infected + new_infections
    states_matrix[,t+1][challenged==1] <- 'C'
    states_matrix[,t+1][susceptible==1] <- 'S'
    states_matrix[,t+1][infected==1] <- 'I'
  }
  return(states_matrix)
}

states_matrix <- fsim2(grid = grid[[1]], N = N[1], sigma, tfinal, amp_a, rate_a, amp_b, rate_b)

dat <- data.frame(ID = grid[[1]]$ID, states_matrix) %>% 
  pivot_longer(cols = starts_with("X"), names_to = 'time', values_to = 'state') %>% 
  mutate(time = as.numeric(gsub("X", '', time)))


#visualize
#for every individual
ggplot(dat, aes(time, ID, fill=state)) +
  geom_tile() +
  scale_fill_manual(values = c('C' = 'orange', 'S' = 'steelblue3', 'I' = 'indianred')) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())

#as a population
dat_pop <- dat %>% count(time, state)
ggplot(dat_pop, aes(time, n, color=state)) +
  geom_line() +
  scale_color_manual(values = c('C' = 'orange', 'S' = 'steelblue3', 'I' = 'indianred'))

```


## Fit model in Stan with more heterogeneity

**Issues: hard to get the sampling started. must supply model with good start values. haven't been able to estimate all of the parameters at once. overfitting probably? need to figure out how to add in replicates. after that, can see how species affects amp_a and amp_b with slopes that vary with species?**

Fitting with no replicates first. 
```{r}
#need to index data by time t AND individual N

#model written in file 'SIR2.stan'
mod2 <- stan_model(file = 'IBM/scripts/model_fitting/SIR2.stan')

# format data for STAN
dat_stan <- list(
  N = as.integer(N[1]),
  t = as.integer(tfinal),
  C = matrix(as.integer(states_matrix=="C"), nrow = N[1]),
  S = matrix(as.integer(states_matrix=="S"), nrow = N[1]),
  I = matrix(as.integer(states_matrix=="I"), nrow = N[1]),
  dist_mat =  dist_matrix,
  sigma_upper = .1
)
str(dat_stan)

initf1 <- function() {
  list(amp_a = .6, amp_b = .1, rate_b = 1, sigma = .005)
}

#run model
fit2 = sampling(mod2,dat_stan, iter = 1500, chains = 1, init = initf1)

#check it out
#amp_a=.6, amp_b=.1, rate_b=1, sigma=.001
precis(fit2, digits = 4) %>% plot #sigma slightly high, amp_b slightly high
stan_dens(fit2)
traceplot(fit2)


amp_a <- .6
rate_a <- .5
amp_b <- .1
rate_b <- 1
tq <- 10
sigma <- .005


#scratch
headmat(dist_matrix^2)
headmat(dist_matrix*dist_matrix)
#how to square a matrix in stan. loop over the vectors. 
newmat <- matrix(NA, nrow=nrow(dist_matrix), ncol=nrow(dist_matrix))
for(i in 1:nrow(dist_matrix)){
  newmat[,i] <- dist_matrix[,i] * dist_matrix[,i]}
headmat(newmat)

headmat(dist_matrix)
headmat((dist_matrix[,1]) %*% dist_matrix)
contact_mat <- f_kernel(dist_matrix, .005)
headmat(contact_mat)
testmat <- (contact_mat %*% dat_stan$I)
headmat(testmat)
melted_cormat <- reshape2::melt(testmat) %>% rename(time="Var2", ID="Var1")
ggplot(data = melted_cormat, aes(x=time, y=ID, fill=value)) + 
  geom_tile()+scale_fill_viridis_c()
```

### stationary alpha/beta, estimate sigma

```{r}

#NEW FUNCTIONS----

#beta will be a product of contact rate X probability of transmission given contact
#contact rate may vary with distance
dist_decay <- function(x, Sigma){
  #equation from Kleczkowski et al. (1997).  sigma scales the rate of the decay. Other parameters in the paper affect a constant, but it cancels out when I standardize it.
  exp(-Sigma*x^2)
}

#probability of transmission given contact varies with time
alpha_curve <- function(t, amp=.4, rate=1) amp*exp(-rate*t)
beta_curve <- function(t, amp=.2, rate=3, tq=11) {
  #amp=amplitude
  #rate=variance of curve. larger is less variance.
  #tq=time of peak
  amp*exp(-rate*(log(t/tq))^2)
}

#will assume different spatial structures. Contact may be equally mixed, be between nearest neighbor only, or decay with distance. 

dist_equal <- function(grid) matrix(1, nrow = length(grid), ncol = length(grid))#homogenously mixed, contacts all same
dist_NN <- function(grid) matrix(as.numeric(rgeos::gTouches(grid, byid = T)), nrow = length(grid), ncol = length(grid)) #nearest neighbor
dist_euc <- function(grid) as.matrix(dist(cbind(grid$x, grid$y)))*10 #euclidian distance
f_kernel <- function(dist_matrix, sigma){
  contact_rt <- dist_decay(dist_matrix, Sigma = sigma)
  diag(contact_rt) <- 0 #Self can't infect self.
  return(contact_rt)
}

#that introduces new parameters: amp_a, rate_a, amp_b, rate_b, tq, sigma. 
#When I model different species, I'm going to assume that species change just the amplitude. Possibly a restrictive assumption, but can't introduce too many parameters. 


#PARAMETERS----

#same parameters as above
pinoc <- .05 #percent inoculated
s <- 1 #number of spp
spp <- "sp1"
tfinal <- 20 #time steps
grid <- sample_community(1, pinoc, 1) 
N <- length(grid)

#new parameters (will be estimated in model fitting)
alpha <- .6
beta <- .05
sigma <- .005
dist_matrix <- dist_euc(grid)
contact_mat <- f_kernel(dist_matrix, sigma = sigma) #contact matrix. for NN and mass action, contact matrix will be the same as the dist_matrix.

#keep track of all individuals' states
agents <- grid@data
states_matrix <- matrix(NA, nrow=N, ncol = tfinal)
states_matrix[,1] <- as.character(agents$state0)

#vectors denoting specific states
challenged <- as.numeric(agents$state0=='C')
susceptible <- 1-challenged
infected <- rep(0, N)


############################################################
#simulate
############################################################

#Run the simulation
for(t in 1:(tfinal-1)){ 

  #calculate probabilities of infections
  P_ci <- (1 - exp(-(alpha + beta*(contact_mat %*% infected))))*challenged
  P_si <- (1 - exp(-beta*(contact_mat %*% infected)))*susceptible
  
  #generate new infections
  CI <- rbinom(N, 1, P_ci)
  SI <- rbinom(N, 1, P_si)
  new_infections <- CI + SI
  
  #update states
  challenged <- challenged - CI
  susceptible <- susceptible - SI
  infected <- infected + new_infections
  states_matrix[,t+1][challenged==1] <- 'C'
  states_matrix[,t+1][susceptible==1] <- 'S'
  states_matrix[,t+1][infected==1] <- 'I'
}

dat <- data.frame(ID = agents$ID, states_matrix) %>% 
  pivot_longer(cols = starts_with("X"), names_to = 'time', values_to = 'state') %>% 
  mutate(time = as.numeric(gsub("X", '', time)))


#visualize
#for every individual
ggplot(dat, aes(time, ID, fill=state)) +
  geom_tile() +
  scale_fill_manual(values = c('C' = 'orange', 'S' = 'steelblue3', 'I' = 'indianred')) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())

#as a population
dat_pop <- dat %>% count(time, state)
ggplot(dat_pop, aes(time, n, color=state)) +
  geom_line() +
  scale_color_manual(values = c('C' = 'orange', 'S' = 'steelblue3', 'I' = 'indianred'))

# format data for STAN
dat_stan <- list(
  N = as.integer(N),
  t = as.integer(tfinal),
  C = matrix(as.integer(states_matrix=="C"), nrow = N),
  S = matrix(as.integer(states_matrix=="S"), nrow = N),
  I = matrix(as.integer(states_matrix=="I"), nrow = N),
  dist_mat =  dist_matrix,
  sigma_upper = .1
)

#model written in file 'SIR2-2.stan'
mod2_2 <- stan_model(file = 'IBM/scripts/model_fitting/SIR2-2.stan')

#run model
fit2_2 = sampling(mod2_2, dat_stan, iter = 2000, chains = 2, init=.01)

#check it out
stan_dens(fit2_2)
traceplot(fit2_2)
precis(fit2_2)

#alpha=.6, beta=.05, sigma=.005
```





