---
title: "Plant SIR model"
author: "Lisa Rosenthal"
date: "3/31/2020"
output:
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(rstan)
require(rethinking)
```

Wes is helping me figure out the best way to parameterize my disease model to the mesocosm data. Much of what is written below is copied from Wes. Main differences are that I changed the states from absolute numbers to densities, which should solve the problem of C->C becoming negative.  

## The model  

It's a discrete time model that describes the transition probilities from three possible states, Challenged, Susceptible, and Infected. C + S + I = N number of individuals in the community. For now, we'll ignore space and species differences. We can  model the transition probabilities between states at $t$ to $t+1$:       

|   | S  |  C | I  |
|---|---|---|---|
| S  | $1-(1-e^{-\beta \frac{I}{N}})$  | 0 |  $1-e^{-\beta \frac{I}{N}}$ |
| C  | $1-e^{-\delta}$  |$1 - (1-e^{-\delta})-(2-e^{-\alpha}-e^{-\beta \frac{I}{N}})$ | $2-e^{-\alpha}-e^{-\beta \frac{I}{N}}$|
| I  | 0  | 0  | 1  |

## Turning observations into data  

Each plant's transition between time steps constitutes an observation. So if there are seven time steps and two plants, there are twelve total observations. Let’s say the state of plant 1 evolves in time like S -> S -> S -> I-> I -> I -> I and plant 2 evolves like C -> C -> I -> I -> I -> I -> I. I would think of each arrow as an observation, where you look up the probability from the transition probability matrix where the state to the left of the arrow is the row index and the state to the arrow’s right is the column index.  

*Wes shows a series of matrices for a given plant and time showing which state change happened. I'm not going to write out all of them.*  

I think it is natural to “unwrap” these matrices into vectors, so the $t = 0$ step representing S -> S would look like:  
$y_0 = (1, 0, 0, 0, 0, 0, 0, 0, 0)$  

This is an observation of a multinomial random variable where the probabilities were those of the first row of
the transition probability matrix (and zeroes elsewhere because there is zero probabilit of going, for instance,
C -> I when the starting state was S.) Thus, the probability governing the multinomial random varaible y0 is:
$p_0 = (1-(1-e^{-\beta \frac{I}{N}})), 1-e^{-\beta \frac{I}{N}}, 0, 0,0,0, 0, 0, 0)$

## Implementing the model

Unfortunately because probability vector $p_t$ changes with time, we can't just model the observations as `y ~ categorical(p)`. Wes decided to break the data into three pieces: one for each starting state (since these correspond to rows of the transition probability matrix.)  

## Code

Here I create the data that were descibed before and estimate parameters of the model. Running the models below generates errors because the probability C -> C is outside of (0, 1). In fact, it is negative. I suspect that one of us has made a math error in the probabilities - probably the error is in the calculation of $P$(C−> I) (my opinion).

```{r}
#transition indicators for subject 1 across the time steps
subj1 <- matrix(c(
  rep(c(1, 0, 0, 0, 0, 0, 0, 0, 0), 2),
  c(0, 0, 1, 0, 0, 0, 0, 0, 0),
  rep(c(0, 0, 0, 0, 0, 0, 0, 0, 1), 3)), 
  nrow=6, ncol=9, byrow = TRUE)

#transition indicators for subject two
subj2 <- matrix( c( 
  c(0, 0, 0, 0, 1, 0, 0, 0, 0),
  c(0, 0, 0, 0, 0, 1, 0, 0, 0),
  rep(c(0, 0, 0, 0, 0, 0, 0, 0, 1), 4)), 
  nrow=6, ncol=9, byrow = TRUE)

#combine the data
raw <- rbind(subj1, subj2)

#change from indicators to labels
dat <- apply(raw, 1, function(col) which(col>0))
dat

#isolate the C, S, and I data
S_obs <- dat[ dat >= 1 & dat <= 3 ] - 0
C_obs <- dat[ dat >= 4 & dat <= 6 ] - 3
I_obs <- dat[ dat >= 7 & dat <= 9 ] - 6

SCI_code <- " data{
int n_S; // number of rows for each starting state
int n_C;
int n_I;

int S[n_S]; // the observed data for each starting state
int C[n_C];
int I[n_I];

int I_count;
}

parameters{
real<lower=0> alpha;
real<lower=0> beta;
real<lower=0> delta;
}

model{
vector[3] pS;
vector[3] pC;
vector[3] pI; 

// prior distributions for the parameters (total guesswork, no thought went into these)
alpha ~ gamma(1, 1);
beta ~ gamma(1, 1);
delta ~ gamma(1, 1); 

// populate the transition probabilities.
// transition probability from state S
pS[1] = 1 - (1 - exp(-beta * I_count));
pS[2] = 0;
pS[3] = 1 - exp(-beta * I_count);

// transition probability from state C
pC[1] = 1 - exp(-delta);
pC[2] = 1 - (1 - exp(-delta)) - (2 - exp(-alpha) - exp(-beta * I_count));
pC[3] = 2 - exp(-alpha) - exp(-beta * I_count);

// transition probability from state I
pI[1] = 0;
pI[2] = 0;
pI[3] = 1;

// define the distribution of the observed data
S ~ categorical( pS );
C ~ categorical( pC );
I ~ categorical( pI );
} "

SCI_model <- stan_model(model_code = SCI_code)

#treating states as absolute numbers
data_list_numbers <- list(
  n_S = length(S_obs),
  n_C = length(C_obs),
  n_I = length(I_obs),
  S = S_obs,
  C = C_obs,
  I = I_obs,
  I_count = 7
)

#Model from Wes doesn't run for me
#myfit_numbers <- sampling( SCI_model, data=data_list_numbers, chains=4 ) 

#Chain 1: Rejecting initial value:
#Chain 1:   Error evaluating the log probability at the initial value.
#Chain 1: Exception: categorical_lpmf: Probabilities parameter is not a valid simplex. Probabilities parameter[2] = -1.29158, but should be greater than or equal to 0  (in 'model15163277fea99_1ec194add31dbadf4c784bfac45dbd18' at line 48)

#Chain 1: 
#Chain 1: Initialization between (-2, 2) failed after 100 attempts. 
#Chain 1:  Try specifying initial values, reducing ranges of constrained values, or reparameterizing the model.
#[1] "Error in sampler$call_sampler(args_list[[i]]) : Initialization failed."

```

I've always learned that for density dependent transmission, $dI/dt = \beta SI$ and for frequency dependent transmission $dI/dt = \beta SI/N$. However, the model that uses absolute numbers of infecteds didn't run, but when I is treated as a proportion, it does run. Definitely a bit confused since I'm not sure how you would represent density dependence when infecteds are in relative terms. Anyways, below is the model that divides by $N$. 

```{r, stanmod2}
#treating states as densities

SCI_code_densities <- " 
data{
// number of rows for each starting state
int n_S; 
int n_C;
int n_I;
// the observed data for each starting state. densities, not integers.
int S[n_S]; 
int C[n_C];
int I[n_I];
int I_count;
int N;
}

parameters{
real<lower=0> alpha;
real<lower=0> beta;
real<lower=0> delta;
}

model{
vector[3] pS;
vector[3] pC;
vector[3] pI; 

// prior distributions for the parameters (total guesswork, no thought went into these)
alpha ~ gamma(1, 1);
beta ~ gamma(1, 1);
delta ~ gamma(1, 1); 

// populate the transition probabilities.
// transition probability from state S
pS[1] = 1 - (1 - exp(-beta * I_count/N));
pS[2] = 0;
pS[3] = 1 - exp(-beta * I_count/N);

// transition probability from state C
pC[1] = 1 - exp(-delta);
pC[2] = 1 - (1 - exp(-delta)) - (2 - exp(-alpha) - exp(-beta * I_count/N));
pC[3] = 2 - exp(-alpha) - exp(-beta * I_count/N);

// transition probability from state I
pI[1] = 0;
pI[2] = 0;
pI[3] = 1;

// define the distribution lf the observed data
S ~ categorical( pS );
C ~ categorical( pC );
I ~ categorical( pI );
} "

SCI_model_densities <- stan_model(model_code = SCI_code_densities)

data_list_densities <- list(
  n_S = length(S_obs),
  n_C = length(C_obs),
  n_I = length(I_obs),
  S = S_obs,
  C = C_obs,
  I = I_obs,
  I_count = 7,
  N=12
)

myfit_densities <- sampling( SCI_model_densities, data=data_list_densities, chains=4 )

precis(myfit_densities, 2)
traceplot(myfit_densities)
plot(myfit_densities)
```








