---
title: "Questioning discrete equations"
author: "Lisa Rosenthal"
date: "4/1/2020"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I have 3 states of plants: **C**hallenged with inoculum, **S**usceptibles, and **I**nfecteds. Since it's a density dependent disease, C, S, and I are the number of individuals within those states. Let's assume $$\frac{dC}{dt} = -\delta C -\alpha C - \beta CI$$ where $\delta$ is the rate that the inoculum decays, $\alpha$ is the inoculum-plant tranmission rate, and $\beta$ is the plant-plant transmission rate.  

With your help, I disretized these equations and came up with the following transition probabilities:  

$$
\begin{aligned}
P_{C\rightarrow S} &= 1 - e^{-\delta} \\
P_{C\rightarrow I} &= 1 - (e^{-\alpha} + e^{-\beta I}) \\
P_{C\rightarrow C} &= 1 - P_{C\rightarrow S} - P_{C\rightarrow I} \\
\end{aligned}
$$

If we assume that $\delta=\alpha=\beta=1$ and that there are 10 infecteds at the time, then CC ends up negative. I'm not sure if the equation for CI, CC, or all are to blame. 
```{r}
delta=alpha=beta=1
I_count=10
CS <- 1 - exp(-delta)
CI <- 1 - (exp(-alpha) + exp(-beta*I_count))
CC <- 1 - CI - CS
CC+CI+CS #sums to 1
CC #negative probability :(
```

